---
title: "Lab01_presentation"
author: "Scott Brown"
date: "1/10/2022"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
```

## R and R Markdown

- Tools for reproducible data analysis
- reproducible, adj. - Capable of being reproduced at a different time or place and by different people.

## Reproducible Data Analysis

- The point of Statistics is providing evidence
- What if all of your evidence involves a heaping load of "take my word on it"?
- Prevent this by providing the tools to reproduce your evidence along with your evidence!

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

## Examples

- [Rpubs](https://rpubs.com/) is a really great place to find examples.
- Unfortunately, Rpubs doesn't also publish the source, which misses the point...
  - [Research-quality article](https://rpubs.com/Ian_W_Richardson/Twitter_Irish_Politics)
  - [Could use a better story](https://rpubs.com/cxh190001/854362)
  - [Also needs better story](https://rpubs.com/silasselfe/Moneyball)
  - [Nice article](https://rpubs.com/toldham2/babynames)
  - [Useful story](https://rpubs.com/Loudc951/848176)
  - [Nice article](https://rpubs.com/samanthallee/facebook-misinformation)
  - [Similar to our Final Project](https://rpubs.com/Felicia_Irawan/848558)
  - [Similar to our Final Project](https://rpubs.com/Ethan_Kostynick/849754)
  
## Today

- Install R, Install RStudio, Install RMarkdown
- Compile an RMarkdown file to html
- Poke around an RMarkdown file
- Take some time to contemplate what parts of the file generate what parts of the document and why!
- Get on Canvas and follow the instruction for Lab 01
- ASK US QUESTIONS!
- One of the TAs will be available for remote learners

