---
output:
  html_document: default
  pdf_document: default
---


# Lab 3: Elementary Probability in R


You will do all your work for this lab in the `console`; so you won't need to touch the `.Rmd` file for this lab.

We'll use the built-in `Titanic` dataset for several questions in this lab

```{r,echo = FALSE}
library(knitr)
```


Load in the `tidyverse` library. Also, run `install.packages("janitor")` and then `library(janitor)`. In your console, type `Titanic`. The `Titanic` dataset displays. To view the dataset, run `view(Titanic)` in the console. You can also learn more about this dataset by running `?Titanic` in the console


First, lets get some practice making contingency tables using `dplyr`. To do so, we'll need to make sure `Titanic` is a `data.frame` object. Run `class(Titanic)` in the console; notice it is a `table` object. Convert it to a `data.frame` object by putting `Titanic = data.frame(Titanic)` in the console, and then running that line. We also recommend setting `N = sum(Titanic$Freq)` in the console. This will store the total number of individuals on the `Titanic` in the variable `N`.

We'll also do some preprocessing on the `Titanic` dataset to make it easier to use in conjunction with the contingency table functions available in `R`. Run the following line in the `console` to update `Titanic`
```{r,eval = FALSE}
Titanic = as.data.frame(lapply(Titanic,rep,Titanic$Freq))

```

Now run the following line in the console:
```{r,eval=FALSE}
Titanic %>%
  tabyl(Age,Survived) %>%
  adorn_totals(c("row","col")) %>%
  adorn_percentages("all") %>%
  adorn_ns

```
The syntax `x %>% f(.,OTHER_ARGS)` feeds `x` into function `f` in the position where the `.` is located (the first argument). The `%>%` is called the `pipe` in R. So `tabyl` takes as its first argument `Titanic`. In this setting, `taybl` creates a contingency table based on the two variables `Age` and `Survived`. The `adorn_totals` function gives row and column totals respectively. The `adorn_percentages` function will put the marginal percentages next to each count. The `adorn_ns` function ensures that both the counts and percentages display.

The `%>%` is from the `magrittr` library which loads autmoatically when you run `library(tidyverse)`. The `tabyl` function is from the `janitor` package.

If you are ever need help using the `tabyl` function, see [here](https://cran.r-project.org/web/packages/janitor/vignettes/tabyls.html). See [here](https://r4ds.had.co.nz/pipes.html) for more information on the `pipe` operator.

## Question 1: Learning Outcome -- Using `tabyl` to Construct a Contingency Table

Write code to produce a 3 x 2 table of `Class` against `Survived`. Run it. Store the newly created data frame as a variable in your working environment. Now answer question 1 on Canvas.


## Question 2: Learning Outcome -- Assessing whether events are independent

Let `A` be the event `Being a Member of 1st Class on Titanic`. Let `B` be the event `Surviving the Titanic`. Are the events `A` and `B` independent?

Hint: Use the table you constructed in question 1

## Question 3: Learning Outcome -- Assessing whether events are mutually exclusive

Let `A` be the event `First class child or second class child`. Let `B` be the event `DIED on the Titanic`. Are the events `A` and `B` mutually exclusive?

Hint: Run the below lines to create a table that `filters` by First or Second class child, and displays the total frequency of Survived, and Died.
```{r,eval = FALSE}
#Assuming you have set Titanic = data.frame(Titanic) in your environment
Titanic %>%
  filter((Class == "1st" | Class == "2nd") & Age == "Child") %>%
  group_by(Survived) %>%
  summarise(n = sum(Freq))

```

## Question 4: Learning Outcome -- Mututal Exclusivity vs Independence

Based on your answer to question 3, are the events `First class child or second class child` and `DIED on the Titanic` independent events?


## Question 5: Learning Outcome -- Using the Addition Rule

Let `A` be the event `Surviving the Titanic` and `B` be the event `Being a member of 3rd class`. Find $P(A \cup B)$.

Hint: The table you created in question 1 may be helpful

## READ

Suppose that in the population that was the 2021 class of CS 3130 students, the event of `X = Getting an A on the First Exam` and `Y = Liking Cilantro` are under study. We collected the corresponding 2x2 table with these two variables for the entire class of 200 students.


`Scott: Maybe we should focus on titanic or cilantro but not both....`


## Question 6: Learning Outcome -- Determining Independence from a Population 2x2 Table
```{r,echo=FALSE}
population = data.frame('Got A' = c(60,40,100), 'Did Not Get A' = c(40,60,100),'Total' = c(100,100,200))
row.names(population) = c('Likes Cilantro','Does Not Like Cilantro','Total')
kable(population)
```

Are `Getting an A on the First exam` and `Liking Cilantro` independent in the population of 2021 CS 3130 students?

## READ

Now suppose Albert had no idea whether `Getting an A on the First exam` and `Liking Cilantro` were independent in the population of 2021 CS 3130 students. He wanted to determine this. 

So he collected information from a simple random sample of 12 students. The data from these students is given below



```{r,echo=FALSE}
sample = data.frame('Got A' = c(4,3,7), 'Did Not Get A' = c(4,1,5),'Total' = c(8,4,12))
row.names(sample) = c('Likes Cilantro','Does Not Like Cilantro','Total')
```

```{r,echo=FALSE}
kable(sample)
```

Let's assume now that we knew apriori that that Albert's sample would have $8$ people that `Liked Cilantro` and $7$ people that `Got an A`. Given this sampling constraint, we'll compute the probability that Albert collected this specific table.

We can a logic similar to that which we used in combinations/permutations and "overcounting" to count up all of the different ways to get this table, and then divided by the factor by which we overcount.

- How many ways can we arrange the 8 folks who Like Cilantro, *taking order into account*?
- How many ways can we arrange the 4 who don't?
- And the 7 who got As?
- And the 5 who did not?

Note that each of these arrangements is a "different choice" in some sense.  So we can use the multiplication rule to combine them.

- How many ways can we arrange the 8 folks who like cilantro, the 4 who don't, the 7 who got As and the 5 who didn't, all taking order into account?

However, some of these arrangements "overcount" w.r.t. our goal of creating a contingency table with this exact layout.  Specifically, when we arranged our cilantro-likers, we counted the various arrangments of the 4 who got "A"s, but when we arranged the folks who got "A"s, we AGAIN counted the arrangements of those same 4.  So, we have overcounted by a factor of "the number of ways to arrange 4 people".  Calculate "the number of ways to arrange" the number of people in each cell of the contingency table above, and divide the answer to Q? by this number.

- How many ways can we get a contingency table with 8 folks who like cilantro, 4 who don't, 7 who got As, 5 who didn't, 4 "A" students who liked cilantro, 3 "A" students who didn't, 4 non-"A" students who liked cilantro, and 1 non-"A" student who didn't?

- How many ways can we order 12 students?

Divide your answer in Q? by your answer to Q? to get the probability of getting our specific contingency table given that $8$ folks liked cilantro and $7$ got an "A".


The key to computing this probability is to note that that there are only two undetermined quantities in the 2x2 table if we know $8$ people `Liked Cilantro` and $7$ people `Got an A`. For example `Got A and Liked Cilantro` and `Got A and Didn't Like Cilantro`.

We basically just have to enumerate the number of ways that $4$ people can `Get an A and Like Cilantro` and also $3$ people can `Get an A and Not Like Cilantro`.

There are $8 \choose 4$ ways to select $4$ of the $8$ people who `liked Cilantro`to get an `A`. That covers the enumeration of `Get A and Like Cilantro`. Now we know that if $4$ people `Get A and Like Cilantro`, then $7-4$ people must `Get A and Not Like Cilantro`, since $7$ people total `Got an A`. And $4$ people in total didn't like cilantro. So $4 \choose 3$ is the number of ways to choose the remaining Cilantro haters who will `Get an A`.

So the number of ways to arrive at this table is
\[
  {8 \choose 4} {4 \choose 3}
\]

Due to independence, each way is equally likely. 

So the probability is
\[
P(X and Y = 4|X+NotX = 8,Y+NotY=4) = \frac{{8 \choose 4} {4 \choose 3}}{\sum_{v=0}^{8} {8 \choose v} {4 \choose 7-v}}
\]

## Question 7: Learning Outcome: Computing the Probability of a Table

Suppose again that there are $8$ people that `Liked Cilantro` and $7$ people that Got an `A` and we recieve the following random sample from the population of CS3130 2021 students:

```{r,echo=FALSE}
sample = data.frame('Got A' = c(3,4,7), 'Did Not Get A' = c(5,0,5),'Total' = c(8,4,12))
row.names(sample) = c('Likes Cilantro','Does Not Like Cilantro','Total')
kable(sample)
```
What is the probability of this table (assuming independence of `Liking Cilantro and Getting an A`)?

