---
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(janitor)
```

# Lab 3: Probability in R

## A Study

Back in 2020, Alice Student took a lot of time and effort to gather data for the entire class comparing whether or not someone got an "A" on the first exam with whether or not they like Cilantro.  The data are in `CS3130Cilantro.csv`.  Load in this data into a variable called `pop2020`.  Although we can use the built-in `R` `table` function to get much of this, we're going to use some additional libraries that are popular in the data-analysis community.  Load the `tidyverse` and `janitor` libraries now (recall the `library` and `install.packages` functions).  Use the code below to construct a 2x2 contingency table over the variables `Got.A` and `Likes.Cilantro`.  The syntax `x %>% f(.,OTHER_ARGS)` feeds `x` into the first argument of function `f`.  The %>% is called the pipe, and is useful for "chaining" operations in `R`.  It is loaded from one of the packages in the `tidyverse` library.

```{r,eval=FALSE}
my_data %>%
  tabyl(Got.A, Likes.Cilantro) %>%
  adorn_totals(c("row","col")) %>%
  adorn_percentages("all") %>%
  adorn_ns() %>%
  View()

```

## Question 1: Determining Independence from a Population 2x2 Table

Are `Getting an A on the First exam` and `Liking Cilantro` independent in the population of 2020 CS 3130 students?

## The probability of a sample (given ?)

It's a new year.  Alice has no idea whether `Getting an A on the First exam` and `Liking Cilantro` are independent in the population of **2021** CS 3130 students. They would like to determine this, but do not have the resources to gather info from the entire population of CS3130 students.

So, they decide that they would like to collect information from a simple random sample of 12 students and arrange the data into a 2x2 table like we just did with the previous year's population data.  Because Alice is a meticulous researcher, she decides to do some probability calculations ahead of time to decide exactly what might convince her that the new population is, in fact, **NOT** independent.  As part of these exercises, they examine the following entirely contrived and arbitrary table:

```{r,echo=FALSE}
sample = data.frame('Got A' = c(3,4,7), 'Did Not Get A' = c(5,0,5),'Total' = c(8,4,12))
row.names(sample) = c('Likes Cilantro','Does Not Like Cilantro','Total')
kable(sample, caption="TABLE 1: contrived example contingency table")
```

The fact that $8$ people `Liked Cilantro` in this table (regardless of whether or not they got an "A") doesn't seem to have anything to do with whether Cilantro and "A"s are independent or not - so, to ease our calculations, we'll only be working with samples that contain $8$ people who like cilantro (and similarly for the other numbers in the "margins").

One of the main results that we are interested in is "If the events 'got an A' and 'likes cilantro' are independent in the population, what is the probability of getting a sample that gives this table (given these margins)?".  We will compute this and some further results in small steps over the next several questions.

## Counting, counting, counting

In order to tackle this problem, we can use counting logic similar to that which we used in combinations/permutations and "overcounting" to count up all of the different ways to get this table and then some, and then divide by the factor by which we overcount.  So, we'll just start by counting lots of things.  Compute the following:

- How many ways can we arrange the 8 folks who Like Cilantro, *taking order into account*?
- How many ways can we arrange the 4 who don't?
- And the 7 who got As?
- And the 5 who did not?

You might find the `?factorial` function in `R` to be helpful.  There is also a related `?choose` function.

## Question 2: Counting Practice 1

Taking order into consideration:

How many ways can we order the 8 folks who like cilantro?

## The multiplication rule

Note that each of these arrangements is a "different choice" in some sense.  So we can use the multiplication rule to combine them.

## Question 3: Counting Practice 2

Imagining for a moment that the 8 cilantro haters, 4 cilantro likers, 7 A students and 5 not A students **are all different people**, how many ways are there to order all of these groups? 

## Overcounting

Using the multiplication rule in question 8 leads to "overcounting" certain arrangements that could result in the contingency table displayed above because they are not, in fact, all different people.  For example, some of the cilantro-likers are "A" students. Specifically, when we arranged our cilantro-likers, we counted the various arrangments of the 3 who got "A"s, but when we arranged the folks who got "A"s, we AGAIN counted the arrangements of those same 3.  So, we have overcounted by a factor of "the number of ways to arrange 3 people".   

## Question 4: Counting Practice 3

Calculate "the number of ways to arrange" the people in the cells of the contingency table above, given the layout of the cells.

Hint: Use the same reasoning from Q8, but with the values in the "cells" of the table instead of the margins.

## From counting to probability

Dividing the correct answer to question $3$ by the correct answer to question $4$ tells you the number of ways to get a contingency table with 3 "A" students who liked cilantro, 4 "A" students who didn't, 5 non-"A" students who liked cilantro, and 0 non-"A" students who didn't when we already know there are 8 folks who like cilantro, 4 who don't, 7 who got As, and 5 who didn't.

Now that we've counted up the number of ways that a sample might have resulted in this table, we would like to start computing probabilities.  In order to turn our counting problem into a probability problem, we *need to establish a scenario* where sampling events are "equally likely".

Note from the 2020 population data that the distribution of people who got "A"s, and the distribution of people who "like cilantro" may not be 50/50.  These are the values and ratios "in the margins".  So, getting a sample with a great number of "A"s is less likely than getting a sample with something like the expected proportion of "A"s.  We will deal with this situation by "fixing the margins" and only considering samples that have gotten precisely the number of "A"s, precisely the number of cilantro-likers, etc as the above table.

Lastly, even with the margins fixed, we are not necessarily equally likely to get someone in the various cells, because "got an A" and "likes cilantro" might be related.  So getting someone who "got an A AND likes cilantro" is not as likely as getting someone who "got an A AND doesn't like cilantro" because "likes cilantro" affects the probability that someone got an A.

However, recall our original question: "If the events 'got an A' and 'likes cilantro' are independent in the population, what is the probability of getting a sample that gives this table?"

If the events 'got an A' and 'likes cilantro' are independent, then whether or not someone likes cilantro does NOT affect the probability that someone got an A!

To sum up, by considering only tables with these margins, we can create a scenario where sampling events are "equally likely" by prefacing our results with "If the events 'got an A' and 'likes cilantro' are independent".  Having to preface your results with this message seems like an unfortunate thing to give up at this point, but it will be super important later, so stay tuned.

## Question 5: Probability Practice

Now that we've counted up the number of ways to order individuals in a sample that results in the above table, and we've established conditions under which the outcomes under our probability model are all "equally likely", compute the probability of getting the above table (given fixed margins and independence).

Hint: How many ways can we order *everyone* in the sample?

## The addition rule

Suppose still that there are $8$ people that `Liked Cilantro` and $7$ people that `Got an A`. There are actually only a couple of possible tables in these circumstances. This is because if we know these margins and specify any given cell, then the table is determined (i.e. you can compute the rest of the table). And based on the margins, the upper left cell can definitely only take on the values $0,1,2,3,4,5,6,7$ (and actually less as we'll see). You've already computed the probability when the upper left cell is $6$.

Denoting $P(Table = i | margins, independence)$ as the conditional probability the upper left cell is $i$ given these margins and given that `Getting A` and `Liking Cilantro` are independent of one another, we can compute for $i=0,1,2,3,4,5,6,7$ $P(Table = i| margins)$. The probabilities are as follows:

```{r,echo=FALSE}
vec = 0:7
computeProb <- function(i) {
  return(choose(8,i)*choose(4,7-i))
}
probs = sapply(vec,computeProb)
probs = probs/sum(probs)
frame = data.frame(i = 0:7,prob = probs)
kable(frame, caption="TABLE 2: all possible contingency tables")
```

These probabilities were computed just like you did for the $i = 3$ case.  Check your answer from Question 10 now.

We can compute the probability of getting one table OR another table using the addition rule.

## Question 6: Addition Rule

What is the probability that the cell in the top left is 0 OR 1 OR 2 OR 3 OR 4 OR 5 OR 6 OR 7?

## Using these as evidence

Now recall Alice from a while back.  They are still interested in whether these variables are independent *in the population*.

Toward this end, she notes that given these margins, there is precisely one theoretical table in which "liking cilantro" and "getting an A" *in our sample* are independent.  You can actually solve for this with a little algebra by setting an arbitrary cell to $x$, and solving for $x$ in $P("A" | Cilantro ) = P("A")$ with our given margins.  You can try this on your own if you like.

```{r,echo=FALSE}
sample = data.frame('Got A' = c(8-10/3,7-(8-10/3),7), 'Did Not Get A' = c(10/3,5-10/3,5),'Total' = c(8,4,12))
row.names(sample) = c('Likes Cilantro','Does Not Like Cilantro','Total')
kable(sample, caption="TABLE 3: independent sample")
```

Of course, it is impossible to actually get a sample that gives this table, but this isn't what Alice is interested in.  Alice notes that although there is precisely one table for which the variables in our sample are independent, there are many in which the variables are *dependent*, and there are *grades* of dependence.  If we get a sample for which the top left cell is 5, then $P("A" | Cilantro)$ is closer to $P("A")$ than when the top left cell is 6 and they are even farther if the top left cell is 7.

Based on this notion of "grades of dependence", she decides that getting a sample that is "very dependent" is good evidence that the population is actually not independent.  Because of her previous investigations with probability, she decides that she might have a means to *measure* the strength of said evidence in terms of a probability calculation.  

Alice notices that getting a sample table with these margins "as dependent or even more dependent" as the one they were examining above has low probability exactly "If 'getting an A' and 'liking cilantro' are independent"!  Because of this, she decides that getting a sample table with these margins "as dependent or even more dependent" than the one above is "strong evidence" that our variables are, in fact, NOT independent. 

## Question 7: The addition rule

Identify which of the possible tables from the "all possible tables" figure above are "as or more dependent than" the contrived sample, where "more dependent" is measured as how far the top left cell is from the "independent" value of $4\frac{2}{3}$ (Note that this could be in either direction!).  What is the probability of getting a sample table given these margins and independence that is "as or more dependent than" the contrived sample?

## Performing an experiment

Alice realizes that her margins might not be as above, so she performs similar probability computations and decides on standards of evidence for all of these various scenarios.  We will spare you these numerous computations.  :)

Now that Alice has decided on what she wants to study, how to go about her experiment, what she will consider strong evidence and why, she decides that she is ready to go gather a sample.

She collects a sample, and as luck would have it the margins just happen to coincide with the contrived sample we've been analyzing in this lab.

```{r,echo=FALSE}
sample = data.frame('Got A' = c(7,0,7), 'Did Not Get A' = c(1,4,5),'Total' = c(8,4,12))
row.names(sample) = c('Likes Cilantro','Does Not Like Cilantro','Total')
kable(sample, caption="TABLE 4: collected data")
```

## Question 8: Quality of evidence

Does Alice decide that she has "strong evidence" that "getting an A" and "likes cilantro" are NOT independent in the 2021 *population*?
